---
title: "RMS Titanic"
author: "Isabella Batalla Salinas"
date: Sept.5 2026
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------------|------------------------|------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

The variables in this data frame are class, sex, age, survived, and n.

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

- Write your observations here
- Are there any differences? There is a difference between the total p- If yes, what might account for those differences?

Yes, there is a difference. On Wikipedia, the total count is 2208 while on the data set it's 2201. These differences may be due to some passengers buying a ticket but not boarding, so the records show that they were there; some people may have snuck in or gone under a different name; or simply a miscount due to lost records.

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`

df_titanic %>%
  filter(Survived == "Yes") %>%
  ggplot(
    aes(x = Class, y = n, fill = Sex)) +
      geom_col(position = "dodge")
  

```

**Observations**:

Among survivors, there were more females than males in 1st and 2nd class, and roughly similar counts in 3rd class. Among Crew, however, male survivors far outnumber female survivors. This is likely because the crew was predominantly male, so this doesn't necessarily indicate a higher survival rate for men, just a larger pool of male crew members to begin with. \# Deeper Look <!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot(aes(x = Class, y = Prop, fill = Sex)) +
  geom_col(position = "dodge")
```

**Observations**:

- Write your observations here.

  Compared to Q3, the Q4 plot gives a clearer picture since it controls for group size. Crew looked like it had way more survivors in Q3, but that's just because there were so many crew members overall. Similarly, 3rd class had a large number of passengers, so its survivor counts in Q3 could look misleadingly high or low depending on the group. Switching to proportions makes the comparisons across classes and sexes more fair, since it's showing how likely someone in that group was to survive, not just how many people happened to.

- Is there anything *fishy* going on in your plot?

  Based on the plot alone, it seems like most of the females in 1st and 2nd class survived in proportion to the total amount of females in each class. Males, on the other hand, had a noticeably lower survival proportion in those same classes. One thing to note is that the rows which contain zero such as "Child" and "Crew" aren't on the plot. This is because there were no child crew members on the Titanic. The gap between male and female survival proportion isn't fishy, though, it lines up with the "women and children first" protocol used during the Titanic evacuation.

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}

df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot(aes(x = Class, y = Prop, fill= Sex)) +
  geom_col(position = "dodge") +
  facet_grid(~ Age)

```

**Observations**:

Plotting survival rate according to proportion and sex, we are able to see how, regardless of age, females had a higher survival rate across all classes (and crew). We are also able to see how all children in first and second class survived, while the children in the third class didn't fare as well.

- If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.

  One of the fishy things seen in q4 was how it seemed like both males and females in the first and second classes had a higher and somewhat equal survival rate compared to other classes. In this plot, however, it can be seen how that doesn't exactly hold up. For males, especially, the survival rate in proportion to age and class is not actually that high. One reason it may seem this way is that the Q4 plot doesn't split by age, so each bar combines children and adults into a single proportion. Since nearly all children survived, that gets blended in with the lower adult survival rate, making the combined bar look higher than the adult rate alone.

  Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
